---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## 1. Loading and preprocessing the data

```{r}
measures <- read.csv("activity.csv", stringsAsFactors=FALSE)
```

## 2. What is mean total number of steps taken per day?

```{r}
q1 <- aggregate(measures$steps, by=list(measures$date), FUN=sum, na.rm=TRUE)
colnames(q1) <- c("ActivityDate", "TotalSteps")
hist(q1$TotalSteps, main=" ", xlab = "Total Steps", col="red")
```

####Prinitng Mean and Median values for toal number of steps
```{r}
print(paste("Mean value of Total Number of Steps Daily =  ", mean(q1$TotalSteps)))
print(paste("Mean value of Total Number of Steps Daily =  ", median(q1$TotalSteps)))
```

## 3. What is the average daily activity pattern?
```{r}
q2 <- aggregate(measures$steps, by=list(measures$interval), FUN=mean, na.rm=TRUE)
colnames(q2) <- c("interval", "AverageSteps")
plot(q2$interval, q2$AverageSteps, type = "l", xlab ="Interval", ylab="Average Steps")
maxIndx <- which(q2$AverageSteps == max(q2$AverageSteps))  
```

####Printing below the interval with Maximum Average Steps
```{r}
print(paste("The interval with Maximum Average Steps = ", q2$interval[maxIndx]))
```


## 4. Imputing missing values   

####Calculating below the number of intervals with missing values 
```{r}
print(paste("Number of Intervals with missing values = ",  sum(is.na(measures$steps))))
```

######Imputing the NA values with the average number of steps for that interval
```{r}
q3 <- merge(measures, q2, by="interval")
q3<- q3[order(q3$date, q3$interval), ]
colIndx <- which(is.na(q3$steps))
q3$steps[colIndx]=q3$AverageSteps[colIndx]
```
Creating a new data frame q4 that is same as original values but has NA substituted with avg no of steps
```{r}
q4 <- q3[c("steps","date","interval")]
head(q4)
```
Plot a histogram of total steps after imputing NA values in the original dataset
```{r}
q5 <- aggregate(q4$steps, by=list(q4$date), FUN=sum, na.rm=TRUE)
colnames(q5) <- c("ActivityDate", "TotalSteps")
hist(q5$TotalSteps, main=" ", xlab = "Total Steps", col="red")
```

####Prinitng Mean and Median values for toal number of steps after imputing NAs
```{r}
print(paste("Mean value of Total Number of Steps Daily =  ", mean(q5$TotalSteps)))
print(paste("Mean value of Total Number of Steps Daily =  ", median(q5$TotalSteps)))  
```

####The mean and median of the data set with imputed values (q5) is more than that of the data set in question 1 above (q1).  



## Are there differences in activity patterns between weekdays and weekends?
